Types for R, Empirically Designed and Evaluated

### 2020-05-06 paper/main.tex
pushing current paper version, with comments. im laying out the narrative surrounding the type design decisions

\usepackage{listings}

@@

@@

\section{Background}

\AT{I think a lot of the background we can pull from an older version of the paper.
I think both are milling around this directory.}

\subsection{The R Programming Language}

\AT{Introduce the language, and give some context about its user base.}

\begin{lstlisting}
hello <-  "world"
paste(c("this", "is", "some", "code"))
\end{lstlisting}

\subsection{R-dyntrace}
\label{sec:r-dyntrace}

\AT{Worth talking a bit about the analysis framework?}

\subsection{The Corpus}

@@

@@

\AT{Talk about Aviral's work on evaluation.}

@@

%% Bibliography
%\bibliography{bibfile}

>>>>

\usepackage{listings, wrapfig}

@@

% Macros for type names from the old paper.
\newcommand{\attr}[2]{\ensuremath{#1_{\mathtt{#2}}}\xspace}
\newcommand{\attrclass}[3]{\ensuremath{#1^{\mathtt{#3}}_{\mathtt{#2}}}\xspace}
\renewcommand{\to}{\ensuremath{\rightarrow}\xspace}
\newcommand{\D}{\ensuremath{\small\vec{\mathtt D}}\xspace} % Double
\newcommand{\I}{\ensuremath{\small\vec{\mathtt I}}\xspace} % Integer
\renewcommand{\C}{\ensuremath{\small\vec{\mathtt C}}\xspace} % Character
\renewcommand{\L}{\ensuremath{\small\vec{\mathtt L}}\xspace} % Logical
\newcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\X}{\ensuremath{\small\vec{\mathtt X}}\xspace} % Complex
\newcommand{\Y}{\ensuremath{\small\vec{\mathtt Y}}\xspace} % Symbol
\newcommand{\sY}{\ensuremath{\small{\mathtt Y}}\xspace} % Symbol
\newcommand{\sS}{\ensuremath{\small{\mathtt S}}\xspace} % S4
\newcommand{\sF}{\ensuremath{\small{\mathtt F}}\xspace} % Closure
\newcommand{\sE}{\ensuremath{\small{\mathtt E}}\xspace} % Env
\renewcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\sN}{\ensuremath{\small{\mathtt N}}\xspace}     % Null
%\renewcommand{\l}{\ensuremath{\small L<?>}\xspace}     % List
\renewcommand{\l}{\ensuremath{\small\underline{\mathtt ?}}\xspace}     % List
\newcommand{\sD}{\ensuremath{\small{\mathtt D}}\xspace} % Double
\newcommand{\sI}{\ensuremath{\small{\mathtt I}}\xspace} % Integer
\newcommand{\sC}{\ensuremath{\small{\mathtt C}}\xspace} % Character
\newcommand{\sL}{\ensuremath{\small{\mathtt L}}\xspace} % Logical
\newcommand{\sX}{\ensuremath{\small{\mathtt X}}\xspace} % Complex
\newcommand{\sR}{\ensuremath{\small{\mathtt R}}\xspace} % Raw
\newcommand{\ANY}{\ensuremath{\small{\mathtt ?}}\xspace}     % Any
\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt{#1}}}\xspace}     % list<T>
\newcommand{\M}[1]{\ensuremath{\attr{\vec{\tt #1}}{mat}}\xspace}     % matrix
\newcommand{\df}{\ensuremath{\attr{\l}{df}}\xspace}     % data.frame

@@

%
%
%
%
%
%
\section{Background}

In this section, we will introduce the R programming language, as well as discuss the dynamic analysis framework we used as part of our initial analysis.

%
%
%
%
\subsection{The R Programming Language}

Over the last decade, the R Project has become a key tool for implementing
sophisticated data analysis algorithms in fields ranging from Computational
Biology~\cite{R05} to Political Science~\cite{R:Keele:2008}. At the heart of
the R project is a \emph{vectorized, dynamic, lazy, functional,
  object-oriented} programming language with a rather unusual combination of
features~\cite{ecoop12} designed to ease learning by non-programmer and
enable rapid development of new statistical methods.  The language, commonly
referred to as R was designed in 1993 by Ross Ihaka and Robert
Gentleman~\cite{R96} as a successor to S~\cite{S88}.  First released in
1995, under a GNU license, R rapidly became the lingua franca for
statistical data analysis. Today, there are over 13,000 R packages available
from repositories such as CRAN and Bioconductor.  With 55 R user groups
world-wide, Smith~\cite{eco11} estimates that there are over 2 million
end-users.

As an introduction to R, consider the code snippet in Fig.~\ref{sample} from
a top-level interaction where the user defines a function \code{normSum}
that accepts vectors of integers, logicals, doubles and complex values and
normalizes the vector with respect to its sum and rounds the results. The
function definition does not require type annotations, and all operations
transparently work on vectors of any length and different types.

\begin{figure}[!hb]{\small
\begin{lstlisting}
> normSum <- function( m )  round( m / sum(m), 2)
> normSum(c(1L,3L,6L))
[1] 0.1 0.3 0.6
> normSum(c(1.1,3.3,6.6))
[1] 0.1 0.3 0.6
> normSum(c(1.6,3.3,6.1))
[1] 0.15 0.30 0.55
> normSum(complex(r=rnorm(3),i=rnorm(3)))
[1] 0.49+0.21i 0.30-0.18i 0.22-0.03i
\end{lstlisting}}
\caption{Sample R code}\label{sample}
\end{figure}

In R, function can be called with named parameters, R support variable
argument lists, and arguments can have default values. Putting all of these
together consider the following declaration:

\begin{lstlisting}
f <- function(x, ..., y=3) x + y
\end{lstlisting}

\noindent
Function \k{f} can be called with a single argument \code{f(3)}, with named
argument \code{f(y=4,x=2)} and with a variable number of arguments,
\code{f(1,2,3,4,y=5)}, all of these calls will return \code{6}.

R has a number of features that are not crucial to the present
discussion. We will mention some of them here for completeness.  In R, data
structures are reference counted and have copy-on-write semantics, thus the
assignment \code{x[12]<-3} results in an update to a copy of \code{x} unless
the reference count on that object is 1.  This semantics gives R a
functional flavor while allowing updating in place within loops (the first
update copies, subsequent updates are performed on the copy). Arguments to
functions are evaluated only when needed, they are bundled in so-called
promises which package the original expression (as an abstract syntax tree, or AST), its environment
as well as the result of evaluating the expression. Promises can be
leveraged for meta-programming as it is possible to retrieve the text of a
promise and evaluate it in a different environment.

\subsubsection{Types of Data}

Before attempting to define a type system for R, we should understand the
different kinds of values that programs operate on.  As we will see
different notions of type may emerge depending on how granular we want to
be.

\renewcommand{\k}[1]{{\tt #1}\xspace}

R has one builtin notion of type that can be queried by the \k{typeof}
function. Over the years, programmers have found the need for a richer type
structure and have added {\it attributes}. The best way to think of attributes is
as an optional map from name to values that can be attached to any object.
Attributes are used to encode various type structures. They can be queried
with functions such as \k{attributes} and \k{class}.

\begin{wrapfigure}{r}{6.1cm}
\footnotesize\begin{tabular}{l|c|l@{}}\hline
\multicolumn{3}{l}{\bf Vectorized data types:}  \\\hline
\k{logical}   & \L & vector of boolean values\\
\k{integer}   & \I & vector of 32 bit integer values\\
\k{double}    & \D & vector of 64 bit floating points\\
\k{complex}   & \X & vector of complex values\\
\k{character} & \C & vector of strings values\\
\k{raw}       & \R & vector of bytes\\
\k{list}      & \l & vector of values of any type\\\hline
\multicolumn{3}{l}{\bf Scalar data types:}\\\hline
\k{NULL}      & \sN &  singleton null value\\
\k{S4}        & \sS &  instance of a S4 class \\
\k{closure}   & \sF & a function with its environment\\
\k{environment}&\sE &  a mapping from symbol to value \\\hline
\multicolumn{3}{l}{\bf Implementation data types:}\\\hline
\multicolumn{3}{l}{\k{special},
\k{builtin},
\k{symbol} (\sY),
\k{pairlist},
\k{promise}}\\
\multicolumn{3}{l}{
\k{language},
\k{char},
\k{...},
\k{any},
\k{expression},
}\\
\multicolumn{3}{l}{
\k{externalprt},
\k{bytecode},
\k{weakref}}\\\hline
\end{tabular}\caption{Builtin Types}\label{types}\end{wrapfigure}

Figure~\ref{types} lists all of the builtin types that are provided by the
language. They are the possible return values of \k{typeof}. There is no
intrinsic notion of subtyping in R. But, in many context a \k{logical} will
convert to \k{integer}, and an \k{integer} will convert to \k{double}.  Some
off conversion can occur in corner cases, such as \k{1<"2"} holds and
\k{c(1,2)[1.6]} returns the first element of the vector, as the double is
converted to an integer. R does not distinguish between scalars and vectors
(they are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
  typeof(c(5,5))} \code{ == "double"}. Finally all vectorized data types have a
distinguished missing value denoted by \code{NA}. The default type of
\code{NA} is \k{logical}. We can see that \code{typeof(NA)=="logical"}, but
NA inhabits every type: \code{typeof(c(1,NA)[2])=="double"}.

With one exception all vectorized data types are monomorphic, the exception
is the \k{list} type which can hold values of any other type including
\k{list}. For all monomorphic data types, attempting to store a value of a
different type will cause a conversion. Either the value is converted to the
type of vector, or the vector is converted to the type of the value.

Scalar data types include the distinguished \k{NULL} value, which is also of
type \k{NULL}, instance of classes written using the S4 object system,
closures and environments.  The implementation of R has a number of other
types that are mostly not used by user code, they are listed in
Figure~\ref{types} for reference.

The addition of attributes lets programmers extend the set of types by
tagging data with user-defined attributes. For example, one could define a
vector of four values, \code{x<-c(1,2,3,4)} and then attach the attribute
\k{dim} with a pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From
that point, arithmetic functions will treat \k{x} as a 2x2 matrix. Another
attribute that can be set is the \k{class}.  This attribute can be bound to
a list of class names. For instance, \code{class(x)<-"human"}, set the class
of \k{x} to be \k{human}.  Attributes are thus used for object-oriented
programming. The S3 object system support single dispatch on the class of
the first argument of a function, whereas the S4 object system allows
multiple dispatch (on all arguments). Some of the most widely used data
type, such as data frames, leverage attributes. A data frame, for instance,
is a list of vectors with a class and a column name attribute.

\paragraph{Summary.} The most common values in R computations are vectorized
types. R programs do not have a way to constrain values to be scalar.
\k{NULL} is sometimes used to represent the case when no value is
available. \k{NA} is used within vector to represent missing observations.
Attributes can decorate values and are used as building blocks for
object-oriented programming. A potential type system for R could focus only
on the builtin types, if one wanted to strive for simplicity, or it could
try to capture attributes at the risk of increased complexity.

%
%
%
%
\subsection{R-dyntrace}
\label{sec:r-dyntrace}

\AT{Leaving this one for Aviral.}

R-dyntrace is an efficient dynamic analysis framework for R.

%
%
%
%
\subsection{The Corpus}

@@

%
%
\subsubsection{Cross-Validation Against Kaggle}

Kaggle is a website offering an environment for hosting Jupyter notebooks, and importantly provides access to a large number of open-source data sets and R and Python code.
This solved a major issue with R programs hosted on e.g. Github, where often the data is not made available.
Our aim is to run R programs hosted on Kaggle to validate the type signatures we generated for popular R packages.

\AT{Anything else to say on Kaggle, here? We will probably talk more about the corpus we analyzed later.}

@@

\AT{Talk about the evaluation.}

\AT{Ok, so we haven't discussed the story for the results, so I'll just write some disjointed sections, discussing design decisions, and backing them up with the data.
This should give us an idea of what we want to say, and we can decide how to organize everything once we have stuff written.
Some of the things I'll be pulling from the data are: how many functions are higher-order? How many times are structs present? How many times do classes supercede other types?
Basically, we want to use this section to turn the simplicity of our type system from a weakness into a strength. 
The simplicity {\it is} a strength, we just need to make sure that's obvious to the reviewer.}

@@

%% Bibliography
\bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual}
